---
title: "reproducible research - assignment 1"
author: "Guillaume Bourgault"
date: "Saturday, July 18, 2015"
output: html_document
---
```{r setoptions, echo = FALSE, warning = FALSE}
library(knitr)
opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, results = 'hold')
options(scipen = 7, digits = 2)
```
Importing the packages used by the script and setting language to englis
```{r, import_packages}
library(dplyr)
library(reshape2)
library(ggplot2)
library(lubridate)
library(lattice)
library(xtable)
```

```{r set_to_english, echo = FALSE, results = 'hide'}
#my system has French as default language.  It is changed to English
Sys.setlocale('LC_TIME', 'English')
```

The data is first loaded.  
```{r read_data}
filename <- 'activity.csv'
data <- read.csv(filename)
```

##Total steps per day
What is the mean and median of total steps per day?
```{r steps per day}
grouped <- group_by(data, date)
steps_per_day <- summarize(grouped, steps_per_day = sum(steps, na.rm = TRUE))
mean_before_imputation <- mean(steps_per_day$steps_per_day)
median_before_imputation <- median(steps_per_day$steps_per_day)
```
The average total steps per day is `r mean_before_imputation`.  
The median total steps per day is `r median_before_imputation`.  
```{r histogram of steps per day}
hist(steps_per_day$steps_per_day, breaks = 20, 
     main = 'Histogram of steps per day', 
     xlab = 'steps per day')
```

##Average daily activity pattern
Prepare a time series plot of the total of steps per 5 minutes, averaged across all days.
```{r}
grouped <- group_by(data, interval)
pattern <- summarize(grouped, steps_per_5_min = mean(steps, na.rm = TRUE))
plot(pattern$interval, pattern$steps_per_5_min, type = 'l', 
     xlab = 'time', 
     ylab = 'total steps per 5 minutes', 
     main = 'total steps per 5 minutes, averaged across all days')
```

What is the most active moment of the day?
```{r}
most_active <- order(pattern$steps_per_5_min, decreasing = TRUE)[1]
time_average <- as.character(pattern$interval[most_active])
most_active <- order(data$steps, decreasing = TRUE, na.last = TRUE)[1]
time_all <- as.character(data$interval[most_active])
date_all <- as.character(data$date[most_active])
```
When taking the average across all days, the most active moment of the day is `r substr(time_average, 1,1)`:`r substr(time_average, 2,3)`.  
Over the entire dataset, the most active moment of the day is `r substr(time_all, 1,1)`:`r substr(time_all, 2,3)` on `r date_all`.  

##Imputing missing values
The dataset contains many missing values.  These are replaced by the average across all dats for the same 5 minutes interval.  

```{r}
nb_missing_before <- sum(is.na(data$steps))
data = mutate(data, steps_imputated = steps)
for (index in 1:dim(data)[1]) {
  if (is.na(data[index, 'steps_imputated'])) {
    data[index, 'steps_imputated'] <- pattern[pattern$interval==data[index, 'interval'],'steps_per_5_min']
  }
}
nb_missing_after <- sum(is.na(data$steps_imputated))
```
Number of missing values before imputation: `r nb_missing_before`.  
Number of missing values after imputation: `r nb_missing_after`.  

Are the average and median steps changed by the imputation?
```{r}
melted <- data.frame(melt(data, id = c('date', 'interval')))
grouped <- group_by(melted, variable, date)
casted <- data.frame(acast(melted, date ~ variable, sum))
mean_after_imputation <- mean(casted$steps_imputated, na.rm = TRUE)
median_after_imputation <- median(casted$steps_imputated, na.rm = TRUE)
```
The mean changed slightly (from `r mean_before_imputation` to `r mean_after_imputation`).    
The median also changed slightly (from `r median_before_imputation` to `r median_after_imputation`).    
Lets compare the effects of the imputation on the histogram of total steps per day. 
```{r}
for_graph <- summarize(grouped, steps = sum(value, na.rm = TRUE))
q <- qplot(steps, data = for_graph, fill = variable, binwidth = 1000)
q <- q + labs(title = 'total number of steps per day')
print(q)
```

#Week activity VS weekend activity
Is the pattern of steps different during the week and the weekend?
```{r}
data$date <- ymd(data$date)
data <- mutate(data, weekday = weekdays(date))
data <- mutate(data, week_weekend = '')
for (index in 1:dim(data)[1]) {
  if (data[index, 'weekday'] %in% c('Saturday', 'Sunday')) {
    data[index, 'week_weekend'] <- 'weekend'}
  else data[index, 'week_weekend'] <- 'week'
}
grouped <- group_by(data, interval, week_weekend)
pattern <- summarize(grouped, steps_per_5_min = mean(steps, na.rm = TRUE))
pattern$week_weekend <- as.factor(pattern$week_weekend)
xyplot(pattern$steps_per_5_min ~ pattern$interval | pattern$week_weekend, 
       layout = c(1,2), type = 'l', 
       xlab = 'time', ylab = 'steps per 5 minutes')
```

A weekday mornings start earlier and include more steps and a higher peak of steps.  
However, weekdays display less steps during the day.  The number of steps during a weekend is more evenly distributed.  